Reinforcement Learning Methods for Operations Research Applications -- Reloaded

### 2020-12-26 paper/paper.tex
shortened standard discounted RL

This section introduces average reward adjusted reinforcement learning in comparison to the discounted
framework, elaborates on optimiality criteria and provides insights of the underlying algorithm.

@@

In the widely applied discounted framework the value of a state \(V_{\gamma}^{\pol_{\gamma}}(s)\) is
defined as the expected discounted sum of rewards under the stationary policy \(\pol_{\gamma}\) when
starting in state \(s\). Note that the policy \(\pol_{\gamma}\) depends on the selected discount
factor. That is
\begin{align*}
  V_{\gamma}^{\pol_{\gamma}}(s)=\lim_{N \to \infty} E[\sum_{t=0}^{N-1} \gamma^{t} R_{t}^{\pol_{\gamma}}(s)]\tcom
\end{align*}

where \(0 \leqslant \gamma < 1\) is the discount factor and
\(R_{t}^{\pol}(s) = E_{\pol}[ r(s_{t},a_{t}) \mid s_{t} = s, a_{t} =
a]\) % \MS[t]{check if correct!}
the reward received at time \(t\) upon starting in state \(s\) by following policy \(\pol\)
\citep{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults}.

The aim in the discounted framework is to find an optimal policy \(\polopt_{\gamma}\), which when
followed, maximises the state value for all states \(s\) as compared to any other policy
\(\pol_{\gamma}\): \(V_{\gamma}^{\polopt_{\gamma}} - V_{\gamma}^{\pol_{\gamma}} \geqslant 0\). This
criteria is usually referred to as \(\gamma\)-optimality as the discount factor \(\gamma\) is fixed.
However, this also means that the actual value set for \(\gamma\) determines the best achievable
policy. For instance, as can be seen in Figure~\ref{fig:printer} setting \(\gamma\) value to
\(<0.8027\) the printer-loop is preferred over the mail-loop by the agent, although the mail-loop
accumulates more reward over time, i.e.\@ selecting the mail-loop is a sub-optimal choice. Observe
that the average reward received per step equals \(1\) for the printer-loop and \(2\) for the
mail-loop. Thus, the (Blackwell-)optimal policy is to choose the mail-loop. However, if
\(\gamma < 3^{-\frac{1}{5}} \approx 0.8027\) an agent using discounted reinforcement learning
chooses the printer loop.

\begin{figure}[t!]
  \centering
  \input{figures/printer}
  \caption{\label{fig:printer} A MDP with a single action choice in
    state \(1\), i.e.\@ two different deterministic policies
    % , in which the agent can choose between doing the printer-loop or the mail-loop  . The
    % Observe that the average reward received per step equals \(1\) for the printer-loop and \(2\)
    % for the mail-loop. Thus, the (Blackwell-)optimal policy is to choose the mail-loop. However,
    % if \(\gamma < 3^{-\frac{1}{5}} \approx 0.8027\) an agent using discounted reinforcement
    % learning chooses the printer loop.
    % %
    \citep[][adapted]{Mahadevan96_OptimalityCriteriaInReinforcementLearning}.
  }
\end{figure}

One idea behind the \(\gamma\)-parameter of the discounted framework is to be able to balance
short-term rewards (low \(\gamma\)-values) and long-term rewards (high \(\gamma\)-values). However,
what seems to be an advantage rather becomes a disadvantage for most applications. The issue is that
the average reward value is non-linearly increased when \(\gamma\) approaches one. However, in
almost all cases the aim is to perform well over time which in terms of reward means to first
maximise for a policy with highest average reward before more selectively choosing actions. That is,
we are searching for the policies with highest average reward before considering other criteria.
%
Therefore, in almost all RL studies the discount factor is set to a value very close to \(1\), for
instance to \(0.99\)~\citep[e.g.][]{mnih2015human,mnih2016asynchronous,Lillicrap15}.
%
By doing so the above mentioned non-linear relationship leads to the fact that the state value
consists almost solely of the up-scaled average reward term, whereas the bias values can be
neglected. This leads to diluted state values and thus difficulties in distinguishing actions which
impose policies that possess the same average reward.
%
In contrast to that, average reward RL separately assesses these values and according to these
selectively chooses the best action, comparing one after the other.
%
To illustrate the idea reconsider Figure~\ref{fig:printer}. First the agent picks the policy
according to the highest average reward. Thus, all actions are assessed by an average reward of
\(2\) which is inferred from the Mail-loop. This however, makes choosing the Printer-loop
unattractive; i.e. the policy converges to the optimal policy of choosing the Mail loop.

@@

In the common case of unichain MDPs, in which only a single set of recurrent states exists, the
average reward \(\avgrew^{\pol}(s)\) is equal for all states \(s\)
\citep{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults,Puterman94}.
In the sequel we focus on unichain MDPs in this work and thus may simply refer to it as
\(\avgrew^{\pol}\). A policy that maximises the average reward in every state is called
gain-optimal. Gain-optimality is the least selective criteria an average reward reinforcement
learning aims for.

@@

The addends, where \(n=-1,0,\ldots\) denote the coefficients of the Laurent series expansion,
can be reformulated to following constraint problem with expected reward
\(R^{\pol}(s) = E_{\pol}[ r(s, a) ] = \sum_{a \in \mathcal{A}(s)}p(a | s) r(s, a)\) as shown by
\cite{MillerVeinott1969} and \citet[p.346]{Puterman94}:

\begin{align}
  \label{eq:constr1}\avgrew^{\pol}(s) - E[\avgrew^{\pol}(s)] & = 0 &  & \text{for } n = -1 \\
  \label{eq:constr2}\avgrew^{\pol}(s) + V^{\pol}(s) - E[V^{\pol}(s)] & = R^{\pol}(s) && \text{for } n = 0 \\
  \label{eq:constr3}W^{\pol}_{n-1}(s) + W^{\pol}_{n}(s) - E[W^{\pol}_{n}(s)] & = 0 && \text{for } n \geqslant 1, \text{where } W_{0}^{\pol}(s) = V^{\pol}(s)
\end{align}

>>>>

This section briefly reintroduces the most important concepts of average reward adjusted
reinforcement learning, elaborates on optimality criteria and provides insights of the underlying
algorithm. For a more extensive introduction to average reward reinforcement learning we refer to
\citet{schneckenreither2020average}.

@@

In the widely applied discounted framework the value of a state \(V_{\gamma}^{\pol_{\gamma}}(s)\) is
defined as the expected discounted sum of rewards under the stationary policy \(\pol_{\gamma}\) when
starting in state \(s\). Note that the policy \(\pol_{\gamma}\) depends on the selected discount
factor. That is
\begin{align*}
  V_{\gamma}^{\pol_{\gamma}}(s)=\lim_{N \to \infty} \E[\sum_{t=0}^{N-1} \gamma^{t} R_{t}^{\pol_{\gamma}}(s)]\tcom
\end{align*}

where \(0 \leqslant \gamma < 1\) is the discount factor and
\(R_{t}^{\pol}(s) = \E_{\pol}[ r(s_{t},a_{t}) \mid s_{t} = s, a_{t} = a]\) the reward received at
time \(t\) upon starting in state \(s\) by following policy \(\pol\)
\citep{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults}.
\MS[t]{how are the update rules?}

The Bellman Equation defined as
\(V_{\gamma}^{\pol}(s) = \E_{\pol}[r + \gamma V_{\gamma}^{\pol}(s')]\) \citep[see
e.g.][p.70]{sutton1998introduction} provides a way to assess the state values using two
consecutively observed states \(s, s'\) and the observed reward \(r\) returned by the system. When
this formula is used as an update rule it provides an algorithm that converges the state values by
iteratively adapting the state value estimates.
%
There are four major issues with discounted RL, which results in the fact that it is inapplicable
for many problems of operations research.

\begin{enumerate}
\item Standard discounted RL can only infer suboptimal policies as the discount factor is strictly
  less than one, i.e.\@ \(\gamma < 1\).
\item It is very difficult to specify a desired balance between the short-term and long-term
  (average reward) rewards. This results from the fact that the long-term rewards are scaled
  exponentially, where the factor depends on the chosen \(\gamma\) value.
\item Episodic MDPs with an average reward per step that is non-zero cannot be solved, as the
  average reward is ignored in the terminal states.
\item The average reward is independently assessed for each state, even though it is the dominating
  part for all state values and thus is shared between more than one or, for unichain MDPs, even all
  states. This usually leads to an exponentially increased number of learning steps required for
  policies to converge.
\end{enumerate}

% The aim in the discounted framework is to find an optimal policy \(\polopt_{\gamma}\), which when
% followed, maximises the state value for all states \(s\) as compared to any other policy
% \(\pol_{\gamma}\): \(V_{\gamma}^{\polopt_{\gamma}} - V_{\gamma}^{\pol_{\gamma}} \geqslant 0\). This
% criteria is usually referred to as \(\gamma\)-optimality as the discount factor \(\gamma\) is fixed.
% However, this also means that the actual value set for \(\gamma\) determines the best achievable
% policy. For instance, as can be seen in Figure~\ref{fig:printer} setting \(\gamma\) value to
% \(<0.8027\) the printer-loop is preferred over the mail-loop by the agent, although the mail-loop
% accumulates more reward over time, i.e.\@ selecting the mail-loop is a sub-optimal choice. Observe
% that the average reward received per step equals \(1\) for the printer-loop and \(2\) for the
% mail-loop. Thus, the (Blackwell-)optimal policy is to choose the mail-loop. However, if
% \(\gamma < 3^{-\frac{1}{5}} \approx 0.8027\) an agent using discounted reinforcement learning
% chooses the printer loop.

% \begin{figure}[t!]
%   \centering
%   \input{figures/printer}
%   \caption{\label{fig:printer} A MDP with a single action choice in
%     state \(1\), i.e.\@ two different deterministic policies
%     % , in which the agent can choose between doing the printer-loop or the mail-loop  . The
%     % Observe that the average reward received per step equals \(1\) for the printer-loop and \(2\)
%     % for the mail-loop. Thus, the (Blackwell-)optimal policy is to choose the mail-loop. However,
%     % if \(\gamma < 3^{-\frac{1}{5}} \approx 0.8027\) an agent using discounted reinforcement
%     % learning chooses the printer loop.
%     % %
%     \citep[][adapted]{Mahadevan96_OptimalityCriteriaInReinforcementLearning}.
%   }
% \end{figure}

% One idea behind the \(\gamma\)-parameter of the discounted framework is to be able to balance
% short-term rewards (low \(\gamma\)-values) and long-term rewards (high \(\gamma\)-values). However,
% what seems to be an advantage rather becomes a disadvantage for most applications. The issue is that
% the average reward value is non-linearly increased when \(\gamma\) approaches one. However, in
% almost all cases the aim is to perform well over time which in terms of reward means to first
% maximise for a policy with highest average reward before more selectively choosing actions. That is,
% we are searching for the policies with highest average reward before considering other criteria.
% %
% Therefore, in almost all RL studies the discount factor is set to a value very close to \(1\), for
% instance to \(0.99\)~\citep[e.g.][]{mnih2015human,mnih2016asynchronous,Lillicrap15}.
% %
% By doing so the above mentioned non-linear relationship leads to the fact that the state value
% consists almost solely of the up-scaled average reward term, whereas the bias values can be
% neglected. This leads to diluted state values and thus difficulties in distinguishing actions which
% impose policies that possess the same average reward.
% %
% In contrast to that, average reward RL separately assesses these values and according to these
% selectively chooses the best action, comparing one after the other.
% %
% To illustrate the idea reconsider Figure~\ref{fig:printer}. First the agent picks the policy
% according to the highest average reward. Thus, all actions are assessed by an average reward of
% \(2\) which is inferred from the Mail-loop. This however, makes choosing the Printer-loop
% unattractive; i.e. the policy converges to the optimal policy of choosing the Mail loop.

Especially due to the fourth issue standard discounted RL is not well applicable in the operations
research domain. This mainly results from the fact that a continuous assessing of the state values
using costs or profit is often required, which leads to intractable long learning phases with high
exploration rate even for small sized problems. But as high exploration rates produce errors in the
state value estimations \citep{MillerVeinott1969} finding a well working parameterisation is
difficult. Average reward reinforcement learning overcomes these issues by separately assessing
addends of the state values.

@@

In the common case of unichain MDPs, in which only a single set of recurrent states exists, the
average reward \(\avgrew^{\pol}(s)\) is equal for all states \(s\)
\citep{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults,Puterman94}.
In the sequel we focus on unichain MDPs in this work and thus may simply refer to it as
\(\avgrew^{\pol}\). A policy that maximises the average reward in every state is called
gain-optimal. % Gain-optimality is the least selective criteria an average reward reinforcement
% learning aims for.

@@

The addends, where \(n=-1,0,\ldots\) denote the coefficients of the Laurent series expansion,
can be reformulated to following constraint problem with expected reward
\(R^{\pol}(s) = \E_{\pol}[ r(s, a) ] = \sum_{a \in \mathcal{A}(s)}p(a | s) r(s, a)\) as shown by
\cite{MillerVeinott1969} and \citet[p.346]{Puterman94}:

\begin{align}
  \label{eq:constr1}\avgrew^{\pol}(s) - \E[\avgrew^{\pol}(s)] & = 0 &  & \text{for } n = -1 \\
  \label{eq:constr2}\avgrew^{\pol}(s) + V^{\pol}(s) - \E[V^{\pol}(s)] & = R^{\pol}(s) && \text{for } n = 0 \\
  \label{eq:constr3}W^{\pol}_{n-1}(s) + W^{\pol}_{n}(s) - \E[W^{\pol}_{n}(s)] & = 0 && \text{for } n \geqslant 1, \text{where } W_{0}^{\pol}(s) = V^{\pol}(s)
\end{align}
